---
title: "Generate GRCh37 refcoding bed"
author: "Nate Olson"
date: '`r Sys.Date()`'
output:
  bookdown::pdf_document2:
    toc: FALSE
---

# Summary
Generating a functional GRCh37 region stratification bed files for use in
variant benchmarking.

# Loading packages and defining variables
```{r message = FALSE}
library(tidyverse)
ftbl_path <- paste0("ftp://ftp.ncbi.nlm.nih.gov/genomes/refseq",
                    "/vertebrate_mammalian/Homo_sapiens/",
                    "all_assembly_versions/GCF_000001405.25_GRCh37.p13",
                    "/GCF_000001405.25_GRCh37.p13_feature_table.txt.gz")
gff_path <- paste0("ftp://ftp.ncbi.nlm.nih.gov/genomes/refseq",
                  "/vertebrate_mammalian/Homo_sapiens/",
                  "all_assembly_versions/GCF_000001405.25_GRCh37.p13",
                  "/GCF_000001405.25_GRCh37.p13_genomic.gff.gz")
faidx_path <- paste0("ftp://ftp-trace.ncbi.nih.gov/1000genomes/",
                  "ftp/technical/reference/",
                  "phase2_reference_assembly_sequence/hs37d5.fa.gz.fai")
```

# Extracting CDS from RefSeq GFF

## Getting Chromosome Assessions
Using feature table to extract chromosome assessions.
Feature table downloaded from `r ftbl_path`.

```{r}
ftbl_md5 <- "c0587443810e3d62da505fb653c28dd3"

ftbl_file <- "GCF_000001405.25_GRCh37.p13_feature_table.txt.gz"
if (!file.exists(ftbl_file)) {
  download.file(url = ftbl_path, destfile = ftbl_file)

  ## MD5 check
  dwn_md5 <- tools::md5sum(ftbl_file)

  if(ftbl_md5 != dwn_md5){
    warning("MD5 for downloaded feature table does not match expected MD5")
  }
}

## defining column types and names
ftbl_col_names <- read_lines(ftbl_file, n_max = 1) %>%
  str_split(pattern = "\t") %>% unlist()
ftbl_col_types <- "cccccccddcclcccdlddc"

feature_table <- read_tsv(ftbl_file, comment = "#",
                          col_names = ftbl_col_names,
                          col_types = ftbl_col_types)

chrom_accn_df <- feature_table %>%
  dplyr::select(chromosome, genomic_accession, assembly_unit, seq_type) %>%
  distinct()
```


## Downloading and Parsing GFF File
Extracting RefSeq CDS from GFF file.
GFF file downloaded from `r gff_path`.

```{r}
gff_md5 <- "e84f4e0102b6c4f4cc0035b8abf1038f"

gff_file <- "GCF_000001405.25_GRCh37.p13_genomic.gff.gz"
if (!file.exists(gff_file)) {
  download.file(url = gff_path, destfile = gff_file)

  ## MD5 check
  dwn_md5 <- tools::md5sum(gff_file)

  if(gff_md5 != dwn_md5){
    warning("MD5 for downloaded feature table does not match expected MD5")
  }
}

## defining column types and names
gff_table <- read_tsv(gff_file, col_names = c("seqid", "source", "type", "start", "end",
                              "score", "strand", "phase", "attributes"),
                comment = "#")

## Exon table with only RefSeq entries
exon_table <- gff_table %>%
  filter(type == "CDS",
         grepl("RefSeq", source)) %>%
  rename(genomic_accession = seqid) %>%
  left_join(chrom_accn_df)

## Extracting CDS for chromosomes and converting to 3 column table.
exon_table_3col <- exon_table %>%
    filter(chromosome %in% c(1:22, "X", "Y", "MT"),
           seq_type %in% c("chromosome","mitochondrion"),
           assembly_unit %in% c("Primary Assembly", "non-nuclear"))%>%
    select(chromosome, start, end) %>%
    ## Sorting table by chromosome and feature start position
    arrange(chromosome, start) %>%
    mutate(chrom = paste0("chr", chromosome)) %>%
    select(chrom, start, end) %>%
  distinct()

## Write to table as a bed file
tmp_bed <- tempfile(fileext = ".bed")
write_tsv(exon_table_3col, tmp_bed, col_names = FALSE)
```

# Preparing Stratification Files
Generating bed file with merged overlapping CDS regions as well as notin bed.
See mappability documentation for how GRCh38 genome bed file was generated.

```{r}
merged_bed_file <- "GRCh37_refseq_cds_merged.bed"
system2("bedtools", args = c("merge", "-i", tmp_bed),
        stdout = merged_bed_file)

## Generating genome bed for subtractBed
faidx_file <- "hs37d5.fa.gz.fai"
faidx_md5 <- "bb77e60e9a492fd0172e2b11e6c16afd"
genome_bed_file <- "human.hg19.chroms.only.genome.bed"

if (!file.exists(faidx_file)) {
  download.file(url = faidx_path, destfile = faidx_file)

  ## MD5 check
  dwn_md5 <- tools::md5sum(faidx_file)

  if(faidx_md5 != dwn_md5){
   warning("MD5 for downloaded feature table does not match expected MD5")
  }
}


faidx_df <- read_tsv(faidx_file, col_names = c("CHROM","SIZE","X1","X2","X3")) %>% 
  filter(CHROM %in% c(1:22,"X","Y")) %>% 
  mutate(CHROM =paste0("chr", CHROM)) %>% 
  mutate(START = 1) %>% 
  select(CHROM, START, SIZE)

write_tsv(faidx_df, path = genome_bed_file, col_names = FALSE)


## Generating not-in bed
notin_merged_bed_file <- "notin_GRCh37_refseq_cds_merged.bed"
system2("subtractBed", args = c("-a", genome_bed_file, "-b", merged_bed_file),
        stdout = notin_merged_bed_file)

## Compressing stratificaiton beds
system2("bgzip", merged_bed_file)
system2("bgzip", notin_merged_bed_file)
```

# File Sanity Checks

```{r}
merged_bed <- read_tsv(paste0(merged_bed_file, ".gz"),
                       col_names = c("chrom","start","end"))
total_cds <- sum(merged_bed$end - merged_bed$start)

notin_merged_bed <- read_tsv(paste0(notin_merged_bed_file, ".gz"),
                       col_names = c("chrom","start","end"))
notin_total_cds <- sum(notin_merged_bed$end - notin_merged_bed$start)
```


Number of bases in merged `r format(total_cds, big.mark = ",")`.
Number of bases not-in merged `r format(notin_total_cds, big.mark = ",")`.
Total bases (in + not-in) `r format(total_cds + notin_total_cds, big.mark = ",")`.

# System Information

```{r}
s_info <- devtools::session_info()
print(s_info$platform)
s_info$packages %>%
  filter(attached) %>%
  dplyr::select(package, loadedversion, source) %>%
      knitr::kable()
```

## Software Versions
__BEDtools__
```{r}
system("bedtools --version",intern = TRUE)
```

__bgzip__
```{r}
system("bgzip --version",intern = TRUE)
```
